Equalizer\\Programming and User Guide

### 2010-10-01 Developer/ProgrammingGuide/paper.tex
Update Programming Guide with SystemPipe and SystemWindow renaming.
[ ] May break build
[ ] May break existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{14cm}
\begin{textblock}{}(0,4.5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock} {\Large The official reference for developing and
  deploying parallel, scalable OpenGL\texttrademark\ applications based on
  the Equalizer parallel rendering framework}\\\vspace{1cm}
Version 1.7.6 for Equalizer 0.9.1+svn\\[\medskipamount]
%  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
%\clearpage
\listoffigures \vfill
\section*{Revision History}
{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
    1.7.0   & Sep 09, 2009 & Updates based on hardcopy print\\
    1.7.1   & Oct 20, 2009 & Updates for new eq::GLWindow\\
    1.7.2   & Nov 18, 2009 & Add \sref{sSubpixel} and \sref{sFSAA} for Subpixel compounds\\
    1.7.3   & Feb 02, 2010 & Revision for Equalizer 0.9.1\\
    1.7.4   & Feb 22, 2010 & Add \sref{sSlaveCommit}\\
    1.7.5   & Jun 03, 2010 & Add \sref{sClusterSetup}\\
    1.7.6   & Aug 23, 2010 & General update for fabric changes\\
    \hline
  \end{tabularx}}
\clearpage

@@

The \textsf{eq::Window} class encapsulates a drawable and an OpenGL
context. The drawable can be an on-screen window or an off-screen
PBuffer or FrameBuffer Object (FBO). The window uses an
\textsf{eq::OSWindow}, which abstracts and manages
window-system-specific handles to the drawable and context, e.g., an X11
window \textsf{XID} and \textsf{GLXContext} for the glX window system.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/layout.png}
  {\caption{\label{fLayout}A Layout with four Views}}
\end{wrapfigure}

@@

Enable the layout on each desired canvas by adding it to the
canvas. Since canvases reference layouts by name or index, layouts have
to be configured before their respective canvases in the configuration
file.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/umlObject.pdf}
  {\caption{\label{fUMLObject}eq::fabric::Serializable and eq::net::Object}}
\end{wrapfigure}
The creation of distributed, versioned objects is simplified when using
\textsf{eq::fabric::Serializable}, which implements one common way of tracking
data changes in versioned objects. The concept of a dirty bit mask is used to
mark parts of the object for serialization, while preserving the capability to
inherit objects. Other ways of implementing change tracking, e.g., using
incarnation counters, can still be implemented by using \textsf{eq::net::Object}
which leaves all flexibility to the developer. \fig{fUMLObject} shows the
relationship between \textsf{eq::fabric::Serializable} and
\textsf{eq::net::Object}.

@@

@@

In \textsf{eqPly}, the version of the only frame-specific object
\textsf{FrameData} is passed as the per-frame id from
\textsf{Config::startFrame} to the frame task methods. The pipe uses
this version to update its instance of the frame data to the current
version, and then unlocks its child entities by calling
\textsf{startFrame}:

{\footnotesize\begin{lstlisting}
void Pipe::frameStart( const uint32_t frameID, const uint32_t frameNumber )
{
    node->waitFrameStarted( frameNumber );

    _frameData.sync( frameID );
    startFrame( frameNumber );
}
\end{lstlisting}}

@@

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/osWindow.pdf}
  {\caption{\label{fOSWindow}eq::OSWindow UML Class Hierarchy}}
\end{wrapfigure}
The particulars of creating a window and OpenGL context depend on the
window system used. One can either use the implementation provided by
the operating system, e.g., AGL, WGL or glX, or some higher-level
toolkit, e.g., Qt.

All window-system specific functionality is implemented by a
specialization of \textsf{eq::OS\-Window}. The \textsf{OSWindow} class
defines the minimal interface to be implemented for a new window
system. Each \textsf{Window} uses one \textsf{OSWindow} during
execution. This separation allows an easy implementation and adaption to
another window system or application.

Equalizer provides a generic interface and implementation for the three
most common window systems through the OpenGL specific 
\textsf{eq::GL\-Window} class: AGL, WGL and glX. The interfaces define the
minimal functionality needed to reuse other window system specific
classes, for example the AGL, WGL and glX event handlers. The separation 
of the OpenGL calls from the \textsf{OSWindow} permits to implement 
window systems which don't use any OpenGL context and therefore to use a 
different renderer. The implementation derived from these interfaces 
provides a sample implementation which honors all configurable window 
attributes.

@@

The initialization sequence uses multiple, override-able task
methods. The main task method \textsf{configInit} calls first
\textsf{configInitOSWindow}, which creates and initializes the
\textsf{OSWindow} for this window. The \textsf{OSWindow} initialization
code is implementation specific. If the \textsf{OSWindow} was
initialized successfully, \textsf{configInit} calls
\textsf{configInitGL}, which performs the generic OpenGL state
initialization. The default implementation sets up some typical OpenGL
state, e.g., it enables the depth test. Most nontrivial applications
do override this task method.

The \textsf{OSWindow} initialization takes into account various
attributes set in the configuration file. Attributes include the size of
the various frame buffer planes (color, alpha, depth, stencil) as well
as other framebuffer attributes, such as quad-buffered stereo,
doublebuffering, fullscreen mode and window decorations. Some of the
attributes, such as stereo, doublebuffer and stencil can be set to
\textsf{eq::AUTO}, in which case the Equalizer default implementation
will test for their availability and enable them if possible.

@@

The window exit happens in the reverse order of the
initialization. First, \textsf{configExitGL} is called to de-initialize
OpenGL, followed by \textsf{configExitOSWindow} which de-initializes the
drawable and context and deletes the \textsf{OSWindow} allocated in
\textsf{configInitOSWindow}.

@@

Events are received by an event handler. The event handler finds the
\textsf{eq::OSWindow} associated to the event. It then creates a generic
\textsf{Event}, which holds the event data in an independent
format. The original native event and this generic \textsf{Event}
form the \textsf{OSWindowEvent}, which is passed to the concrete
\textsf{OSWindow} for processing.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
  {\caption{\label{fEventProcessing}Event Processing}}
\end{wrapfigure}
The purpose of the \textsf{OSWindow} processing method,
\textsf{processEvent}, is to perform window-system specific event
processing. For example, \textsf{AGLWindow::processEvent} calls
\textsf{aglUpdateContext} whenever a resize event is received. For
further, generic processing, the \textsf{Event} is passed on to
\textsf{Window::processEvent}. This \textsf{Event} no longer contains
the native event.

@@

Later Equalizer versions will introduce \textsf{Pipe::pro\-cess\-Event} and
\textsf{PipeEvent} to communicate pipe-specific events, e.g., monitor
resolution changes. It is also likely that an \textsf{OSPipe}
abstraction, similar to the \textsf{OSWindow}, will be created to
separate the window system specific code from the \textsf{eq::Pipe}.

@@

The main idea behind direct send is to parallelize the costly
recomposition for database (sort-last) decomposition. With each
additional source channel, the amount of pixel data to be composited
grows linearly. When using the simple approach of compositing all frames
on the destination channel, this channel quickly becomes the bottleneck
in the system. Direct send distributes this workload evenly across all
source channels, and thereby keeps the compositing work per channel
constant.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\label{fDirectSend}Direct Send Compositing}}
\end{wrapfigure}
In direct send compositing, each rendering channel is also responsible
for the sort-last composition of one screen-space tile. He receives
the framebuffer pixels for his tile from all the other channels. The
size of one tile decreases linearly with the number of source channels,
which keeps the total amount of pixel data per channel constant.

@@

Readback and assemble operations on frames and images are designed to be
asynchronous. They have a start and finish method for both readback and
assemble to allow the initiation and synchronization of the operation.
Currently, only synchronous readback and assembly using
\textsf{glReadPixels} and \textsf{glDrawPixels} is implemented in the
respective start method of the image. Later versions of Equalizer will
implement asynchronous pixel transfers.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
  {\caption{\label{fAssembly}Hierarchy of Assembly Classes}}
\end{wrapfigure}
The offset of input and output frames characterizes the position of the
frame data relative to the framebuffer, that is, the \textbf{window's}
lower-left corner. For output frames this is the position of the channel
relative to the window.

@@

DB compounds have the advantage of scaling any part of the volume
rendering pipeline: texture and main memory (smaller bricks for each
channel), fill rate (less samples per channel) and IO bandwidth for
time-dependent data (less data per time step and channel). Since the
amount of texture memory needed for each node decreases linearly, they
make it possible to render data sets which are not feasible to
visualize with any other approach.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
For recomposition, the 2D frame buffer contents are blended to form a
seamless picture. For correct blending, the frames are ordered in the
same back-to-front order as the slices used for rendering, and use the
same blending parameters. Simplified, the frame buffer images are
`thick' slices which are `rendered' by writing their content to the
destination frame buffer using the correct order.

@@

Please note that the support for CUDA is only enabled if the CUDA libraries are found
in their default locations. Otherwise please adapt the makefiles accordingly.

@@

Equalizer automatically initializes a CUDA context on each pipe (using the specified
device number) whenever \textsf{hint\_cuda\_GL\_interop} is set. The initialization is 
currently done using the CUDA runtime API in order to support device emulation. 
On the other side, if this hint is not set the OpenGL interop feature is not enabled, and 
furthermore, the GPU which executes CUDA code defaults to device 0.

>>>>

\date{
  \vspace{14cm}
\begin{textblock}{}(0,4.5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock} {\Large The official reference for developing and
  deploying parallel, scalable OpenGL\texttrademark\ applications based on
  the Equalizer parallel rendering framework}\\\vspace{1cm}
Version 1.7.7 for Equalizer 0.9.1+svn\\[\medskipamount]
%  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
%\clearpage
\listoffigures \vfill
\section*{Revision History}
{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
    1.7.0   & Sep 09, 2009 & Updates based on hardcopy print\\
    1.7.1   & Oct 20, 2009 & Updates for new eq::GLWindow\\
    1.7.2   & Nov 18, 2009 & Add \sref{sSubpixel} and \sref{sFSAA} for Subpixel compounds\\
    1.7.3   & Feb 02, 2010 & Revision for Equalizer 0.9.1\\
    1.7.4   & Feb 22, 2010 & Add \sref{sSlaveCommit}\\
    1.7.5   & Jun 03, 2010 & Add \sref{sClusterSetup}\\
    1.7.6   & Aug 23, 2010 & General update for fabric changes\\
    1.7.7   & Oct 01, 2010 & SystemPipe and SystemWindow renaming\\
    \hline
  \end{tabularx}}
\clearpage

@@

The \textsf{eq::Window} class encapsulates a drawable and an OpenGL
context. The drawable can be an on-screen window or an off-screen
PBuffer or FrameBuffer Object (FBO).

@@

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/layout.png}
  {\caption{\label{fLayout}A Layout with four Views}}
\end{wrapfigure}
Enable the layout on each desired canvas by adding it to the
canvas. Since canvases reference layouts by name or index, layouts have
to be configured before their respective canvases in the configuration
file.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/umlObject.pdf}
  {\caption{\label{fUMLObject}fabric::Serializable and net::Object}}
\end{wrapfigure}
The creation of distributed, versioned objects is simplified when using
\textsf{eq::fabric::Serializable}, which implements one common way of tracking
data changes in versioned objects. The concept of a dirty bit mask is used to
mark parts of the object for serialization, while preserving the capability to
inherit objects. Other ways of implementing change tracking, e.g., using
incarnation counters, can still be implemented by using \textsf{eq::net::Object}
which leaves all flexibility to the developer. \fig{fUMLObject} shows the
relationship between \textsf{eq::fabric::Serializable} and
\textsf{eq::net::Object}.

@@

The pipe uses an \textsf{eq::SystemPipe}, which abstracts and manages
window-system-specific code for the GPU, e.g., an X11 \textsf{Display}
connection for the glX pipe system.

@@

In \textsf{eqPly}, the version of the only frame-specific object
\textsf{FrameData} is passed as the per-frame id from
\textsf{Config::startFrame} to the frame task methods. The pipe uses
this version to update its instance of the frame data to the current
version, and unlocks its child entities by calling
\textsf{startFrame}:

{\footnotesize\begin{lstlisting}
void Pipe::frameStart( const uint32_t frameID, const uint32_t frameNumber )
{
    startFrame( frameNumber );
    _frameData.sync( frameID );
}
\end{lstlisting}}

@@

The window uses an \textsf{eq::SystemWindow}, which abstracts and manages
window-system-specific handles to the drawable and context, e.g., an X11 window
\textsf{XID} and \textsf{GLXContext} for the glX window system.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/osWindow.pdf}
  {\caption{\label{fOSWindow}SystemWindow UML Class Hierarchy}}
\end{wrapfigure}
The particulars of creating a window and OpenGL context depend on the
window system used. One can either use the implementation provided by
the operating system, e.g., AGL, WGL or glX, or some higher-level
toolkit, e.g., Qt.

All window-system specific functionality is implemented by a
specialization of \textsf{eq::System\-Window}. The \textsf{SystemWindow} class
defines the minimal interface to be implemented for a new window
system. Each \textsf{Window} uses one \textsf{SystemWindow} during
execution. This separation allows an easy implementation and adaption to
another window system or application.

Equalizer provides a generic interface and implementation for the three
most common window systems through the OpenGL specific 
\textsf{eq::GL\-Window} class: AGL, WGL and glX. The interfaces define the
minimal functionality needed to reuse other window system specific
classes, for example the AGL, WGL and glX event handlers. The separation 
of the OpenGL calls from the \textsf{SystemWindow} permits to implement 
window systems which don't use any OpenGL context and therefore to use a 
different renderer. The implementation derived from these interfaces 
provides a sample implementation which honors all configurable window 
attributes.

@@

The initialization sequence uses multiple, override-able task
methods. The main task method \textsf{configInit} calls first
\textsf{configInitSystemWindow}, which creates and initializes the
\textsf{SystemWindow} for this window. The \textsf{SystemWindow} initialization
code is implementation specific. If the \textsf{SystemWindow} was
initialized successfully, \textsf{configInit} calls
\textsf{configInitGL}, which performs the generic OpenGL state
initialization. The default implementation sets up some typical OpenGL
state, e.g., it enables the depth test. Most nontrivial applications
do override this task method.

The \textsf{SystemWindow} initialization takes into account various
attributes set in the configuration file. Attributes include the size of
the various frame buffer planes (color, alpha, depth, stencil) as well
as other framebuffer attributes, such as quad-buffered stereo,
doublebuffering, fullscreen mode and window decorations. Some of the
attributes, such as stereo, doublebuffer and stencil can be set to
\textsf{eq::AUTO}, in which case the Equalizer default implementation
will test for their availability and enable them if possible.

@@

The window exit happens in the reverse order of the
initialization. First, \textsf{configExitGL} is called to de-initialize
OpenGL, followed by \textsf{configExitSystemWindow} which de-initializes the
drawable and context and deletes the \textsf{SystemWindow} allocated in
\textsf{configInitSystemWindow}.

@@

Events are received by an event handler. The event handler finds the
\textsf{eq::SystemWindow} associated to the event. It then creates a generic
\textsf{Event}, which holds the event data in an independent
format. The original native event and this generic \textsf{Event}
form the \textsf{SystemWindowEvent}, which is passed to the concrete
\textsf{SystemWindow} for processing.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
  {\caption{\label{fEventProcessing}Event Processing}}
\end{wrapfigure}
The purpose of the \textsf{SystemWindow} processing method,
\textsf{processEvent}, is to perform window-system specific event
processing. For example, \textsf{AGLWindow::processEvent} calls
\textsf{aglUpdateContext} whenever a resize event is received. For
further, generic processing, the \textsf{Event} is passed on to
\textsf{Window::processEvent}. This \textsf{Event} no longer contains
the native event.

@@

Later Equalizer versions will introduce \textsf{Pipe::pro\-cess\-Event} and
\textsf{PipeEvent} to communicate pipe-specific events, e.g., monitor
resolution changes.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\label{fDirectSend}Direct Send Compositing}}
\end{wrapfigure}
The main idea behind direct send is to parallelize the costly
recomposition for database (sort-last) decomposition. With each
additional source channel, the amount of pixel data to be composited
grows linearly. When using the simple approach of compositing all frames
on the destination channel, this channel quickly becomes the bottleneck
in the system. Direct send distributes this workload evenly across all
source channels, and thereby keeps the compositing work per channel
constant.

In direct send compositing, each rendering channel is also responsible
for the sort-last composition of one screen-space tile. He receives
the framebuffer pixels for his tile from all the other channels. The
size of one tile decreases linearly with the number of source channels,
which keeps the total amount of pixel data per channel constant.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
  {\caption{\label{fAssembly}Hierarchy of Assembly Classes}}
\end{wrapfigure}
Readback and assemble operations on frames and images are designed to be
asynchronous. They have a start and finish method for both readback and
assemble to allow the initiation and synchronization of the operation.
Currently, only synchronous readback and assembly using
\textsf{glReadPixels} and \textsf{glDrawPixels} is implemented in the
respective start method of the image. Later versions of Equalizer will
implement asynchronous pixel transfers.

The offset of input and output frames characterizes the position of the
frame data relative to the framebuffer, that is, the \textbf{window's}
lower-left corner. For output frames this is the position of the channel
relative to the window.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
DB compounds have the advantage of scaling any part of the volume
rendering pipeline: texture and main memory (smaller bricks for each
channel), fill rate (less samples per channel) and IO bandwidth for
time-dependent data (less data per time step and channel). Since the
amount of texture memory needed for each node decreases linearly, they
make it possible to render data sets which are not feasible to
visualize with any other approach.

For recomposition, the 2D frame buffer contents are blended to form a
seamless picture. For correct blending, the frames are ordered in the
same back-to-front order as the slices used for rendering, and use the
same blending parameters. Simplified, the frame buffer images are
`thick' slices which are `rendered' by writing their content to the
destination frame buffer using the correct order.

@@

Please note that the support for CUDA is only enabled if the CUDA libraries are
found in their default locations. Otherwise please adapt the build system
accordingly.

@@

Equalizer automatically initializes a CUDA context on each pipe (using the
specified device number) whenever \textsf{hint\_cuda\_GL\_interop} is set. The
initialization is currently done using the CUDA runtime API to support device
emulation. On the other side, if this hint is not set the OpenGL interop feature
is not enabled, and furthermore, the GPU which executes CUDA code defaults to
device 0.
